`You Draw It': Implementation of visually fitted trends with \texttt{r2d3}

### 2022-07-06 2022-sdss-you-draw-it-manuscript.tex
added more to the validation study and addressed a few comments above.

\usepackage{natbib}
\bibliographystyle{unsrtnat}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
  pdfkeywords={graphics, user interaction, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

We all use statistical graphics, but how do we know that the graphics we
use are communicating \svp{effectively}? Through experimentation,
graphical testing methods allow researchers to conduct studies
\svp{designed to understand how we perceive graphics and perform graphical tasks}
such as differentiation, prediction, estimation, and extrapolation.
\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}

\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
These charts rely on our} ability to identify and detect trends in data.
Our visual system is naturally built to look for structure and identify
patterns,
\svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
\svp{As shown in @fig:gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations. First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points. Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability. Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}

@@

{\centering \includegraphics{./images/gas-prices-1.pdf}

@@

\caption{Weekly average gas prices in the United States, 2019-June 2022.
Additional mental operations a viewer might perform while looking at the
plot are annotated in grey. US Energy Information Administration (2022)}

@@

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points (Finney 1951, mosteller1981eye). Common
methods of fitting trends by eye involved maneuvering a string, black
thread, or ruler until the fit was suitable, then drawing the line
through the set of points. Recently, Ciccione and Dehaene (2021)
conducted a comprehensive set of studies investigating human ability to
detect trends in graphical representations from a psychophysical
approach.\svp{XXX and found what? Describe the results XXX}

\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}
In 2015, the New York Times introduced an interactive feature, called
`You Draw It' (Aisch, Cox, and Quealy 2015), where readers input their
own assumptions about various \svp{metrics of news interest} and compare
these assumptions to reality. The New York Times team used Data Driven
Documents (D3), \svp{a JavaScript library} that allows readers to
\svp{interact with a chart directly by} drawing a line on their computer
screen with a mouse.
\svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}

\svp{We set out to implement} `You Draw It', adapted from the New York
Times feature, as a way to experimentally assess the patterns we see in
data. Here, we provide technical details of the software development,
utilizing interactive graphics in R(R Core Team 2022). We then share
results from our study which validates `You Draw It' as a method for
graphical testing \svp{on linear trends} and apply an appropriate data
analysis method to the participant data.
\svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}

@@

\svp{The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their covid dashboards. Naturally, this same framework was used}
to create the `You Draw It' \svp{feature}.

A challenge of working with D3 is the environment necessary to display
the graphics and images.
\svp{XXX What environment, why is this a challenge, and how did you overcome it?}

@@

\svp{We leveraged the `r2d3` package [@r2d3pkg] to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script. This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data. While not strictly necessary for the validation test of the You Draw It method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.}

We conducted all data simulation and processing in R and output two data
sets - \emph{point data} and \emph{line data} - containing (x, y)
coordinates corresponding to either a simulated point or fitted value
predicted by a statistical model respectively. Then, the \texttt{r2d3}
package converts the data sets in R to JavaScript Object Notation (JSON)
to be interpreted by JavaScript code included with the \texttt{r2d3}
package. Parameters for aesthetic design choices are defined in a list
of options in the \texttt{r2d3} function call; \texttt{r2d3} passes
these to the generated JavaScript code. For instance, we can specify the
buffer space allowed for the \(x\) and \(y\) axes to avoid users
\svp{anchoring} their lines to the axes limits.

@@

\svp{
In order to make use of the original You Draw It code for perceptual experiments, we first needed to accommodate an additional layer. The original NYT features asked participants to draw on a blank coordinate grid, as shown in @fig-nyt-screenshot.
}

@@

\svp{When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line. This requires that we modify the original JavaScript code to accommodate these additional elements, which we generate using `r2d3`, as described above.}

We defined JavaScript functions to draw the initial plot and set up
drawable points for the user drawn line. Drag events in \texttt{D3.js}
are utilized to observe and react to user input,
\svp{as in the original You Draw It script from the NYT}. \footnote{The
  full source code we developed for the YouDrawIt task is available at
  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}

\svp{One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written. While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.}

@@

\svp{During testing of our modifications to the You Draw It script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence. This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded. While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.}

@@

\svp{One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements. Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.}

@@

Shiny Messages are used to communicate the user interaction between the
JavaScript code and the R environment.
\svp{The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code. Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points. This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.}
Once the user is done drawing the line, we save the results of the drawn
line to a SQLite database, using Shiny to pass data from JavaScript to
R. The connections between each portion of the app are shown in
Figure~\ref{fig-you-draw-it-code-sketch}.

@@

\hypertarget{tool-validation}{%
\subsection{Tool Validation}\label{tool-validation}}

We conducted a study in order to validate `You Draw It' as a method for
graphical testing, comparing results to the less technological method
utilized in Mosteller et al. (1981). Results from our study were
consistent with those found in the previous study; when shown points
following a linear trend, participants tended to fit the slope of the
first principal component over the slope of the ordinary least-squares
regression line (Figure 3). This trend was most prominent when shown
data simulated with larger variances. This study reinforced the
differences between intuitive visual model fitting and statistical model
fitting, providing information about human perception as it relates to
the use of statistical graphics.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/pca-plot}}
\caption{Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.}
\label{pca-plot}
\end{center}
\end{figure}

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

Feedback data from conducted studies were collected and stored in a
database for analysis (Figure 4). Within the collected feedback data, we
knew the simulated data points, the predicted values from the
statistical model, and the predicted values from the user drawn line. In
our initial studies, a unique data set was simulated independently for
each participant. Therefore, we evaluate the accuracy of the user drawn
line by observing the deviation, vertical residuals, between the user
drawn line and the predicted values from the statistical model. Due to
participant variation, we use mixed models to evaluate the vertical
residuals in order to statistically compare visually fitted trends to
actual metrics, simulated data models, or statistical regression
results. In our validation study, we first analyzed the residual trend
using a linear mixed model (LMM), thus constraining the fit to a linear
trend (Figure 5). In addition, we fit a Generalized Additive Mixed Model
(GAMM) to allow for flexibility in the residual trend and to generalized
the method to nonlinear models (Figure 6).

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

After validating the `You Draw It' method, we conducted a study
utilizing the new method to evaluate participants ability to make
forecasts for exponentially increasing data on log and linear scales.
Along with analyzing the feedback data with the GAMM method for
flexibility due to nonlinear data, we used spaghetti plots to visualize
participants forecasts compared to the nonlinear least squares
statistical model and make comparisons between two chart design features
(Figure 6).

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/exponential-yloess-spaghetti-plot-2-1}}
\caption{Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data. Participants drawn lines on the linear scale are shown in blue and the log scale are shown in orange. Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.}
\label{exponential-yloess-spaghetti-plot-2-1}
\end{center}
\end{figure}

\hypertarget{future-work}{%
\section{Future Work}\label{future-work}}

In this work, we implemented and validated `You Draw It' as a new method
to measure the patterns we see in data. We demonstrated the use of
generalized additive models to statistically model participant feedback
data and generalized the analysis method to nonlinear trends. While
technical details of the development process are presented here, we
intend to create an R package designed for easy implementation of `You
Draw It' task plots in order to make this tool accessible to other
researchers. Further investigation is necessary to implement this method
with real data in order to facilitate scientific communication.

>>>>

\usepackage{natbib}
\bibliographystyle{unsrtnat}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
  pdfkeywords={graphics, user interaction, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

We all use statistical graphics, but how do we know that the graphics we
use are communicating effectively? Through experimentation, graphical
testing methods allow researchers to conduct studies designed to
understand how we perceive graphics and perform graphical tasks such as
differentiation, prediction, estimation, and extrapolation. Each of
these levels of interaction with a graph require a different method of
engagement with and use of the information presented in a chart. In this
paper, we describe the adaptation of an old tool for graphical testing
and evaluation, eye-fitting, for use in modern web-applications suitable
for testing statistical graphics. We present an empirical evaluation of
this testing method for linear regression, and briefly discuss an
extension of this method to non-linear applications.

One of the most common charts created is a scatter plot of points over
time; these charts show up regularly in news articles and in scientific
publications alike. These charts rely on our ability to identify and
detect trends in data. Our visual system is naturally built to look for
structure and identify patterns, including patterns and trends over
time; many times we do not even notice this process happening
subconsciously. As shown in Figure~\ref{fig-gas-prices}, a viewer
engaging with a plot of weekly average gas prices over time may perform
several cognitive operations. First, they scan the plot and assesses
points to determine if there are any outliers or otherwise remarkable
points. Then, they may fit a rough mental smooth/trend line to the
points to summarize the useful information and remove variability.
Finally, an interested and engaged viewer may pull in additional
contextual information from long-term memory, seeking to explain
variation in the mental `trend' with supplemental information, such as
COVID lock downs which reduced gasoline demand and the beginning of the
war in Ukraine, which wreaked havoc on the supply and demand for global
energy sources.

@@

{\centering \includegraphics{./images/fig-gas-prices-1.pdf}

@@

\caption{\label{fig-gas-prices}Weekly average gas prices in the United
States, 2019-June 2022. Additional mental operations a viewer might
perform while looking at the plot are annotated in grey. US Energy
Information Administration (2022)}

@@

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points Mosteller et al. (1981). Common methods of
fitting trends by eye involved maneuvering a string, black thread, or
ruler until the fit was suitable, then drawing the line through the set
of points.
\ear{Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.}
Recently, Ciccione and Dehaene (2021) conducted a comprehensive set of
studies investigating human ability to detect trends in graphical
representations from a psychophysical approach.
\ear{This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increased with noise and with number of points. 
This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.}

While psychologists and statisticians have been using eye-fitting
techniques to assess our innate perceptual statistical modeling
abilities, news organizations have used similar techniques in order to
draw readers in and demonstrate the difference between readers'
assumptions and reality. In 2015, the New York Times introduced an
interactive feature, called `You Draw It' (Aisch, Cox, and Quealy 2015),
where readers input their own assumptions about various metrics of news
interest and compare these assumptions to reality. The New York Times
team used Data Driven Documents (D3), a JavaScript library that allows
readers to interact with a chart directly by drawing a line on their
computer screen with a mouse. Despite the somewhat different purpose
behind this feature, the D3 driven method used by the NYT is wonderfully
intuitive, and does not require the assumption of linearity, making it
much more adaptable to testing how viewers perceive and predict when
presented with non-linear data.

We set out to implement `You Draw It', adapted from the New York Times
feature, as a way to experimentally assess the patterns we see in data.
Here, we provide technical details of the software development,
utilizing interactive graphics in R (R Core Team 2022). We then share
results from our study which validates `You Draw It' as a method for
graphical testing on linear trends and apply an appropriate data
analysis method to the participant data. We also briefly demonstrate the
use of the `You Draw It' method and analysis on nonlinear data.

@@

The New York Times uses D3, Data Driven Documents (Bostock, Ogievetsky,
and Heer 2011), to create many of the interactive data graphics that
users rely on, from the famous ``election needle'' in 2016 to their
COVID dashboards. Naturally, this same framework was used to create the
`You Draw It' feature.

A challenge of working with D3 is the environment necessary to display
the graphics and images.
\ear{Typically, a server setup is needed in order to render D3 visuals and show output from JavaScript on the webpage.
This can be a barrier to getting started with D3 for someone unfamiliar with server setup and minimal JavaScript coding background.
Resources, such as @wattenberger-fullstack, help novice users overcome this challenge by walking them step by step through setting up a local server and creating their first plot in D3.
In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.}

@@

We leveraged the \texttt{r2d3} package to take data randomly generated
in R and create a D3 plot that could be modified with the You Draw It
script. This step allowed us to more easily connect the data generation
process with the resulting D3 code without having to generate new code
manually each time we re-generated the data. While not strictly
necessary for the validation test of the `You Draw It' method compared
to past linear regression methods presented in this paper, this
abstraction makes it much easier to test arbitrary or model-generated
data which is unique to each participant.

We conducted all data simulation and processing in R and output two data
sets - \emph{point data} and \emph{line data} - containing (x, y)
coordinates corresponding to either a simulated point or fitted value
predicted by a statistical model respectively. Then, the \texttt{r2d3}
package converts the data sets in R to JavaScript Object Notation (JSON)
to be interpreted by JavaScript code included with the \texttt{r2d3}
package. Parameters for aesthetic design choices are defined in a list
of options in the \texttt{r2d3} function call; \texttt{r2d3} passes
these to the generated JavaScript code. For instance, we can specify the
buffer space allowed for the \(x\) and \(y\) axes to avoid users
anchoring their lines to the axes limits.

@@

In order to make use of the original `You Draw It' code for perceptual
experiments, we first needed to accommodate an additional layer. The
original NYT features asked participants to draw on a blank coordinate
grid, as shown in Figure~\ref{fig-nyt-screenshot}.

@@

When testing perception of graphics, however, we want to pre-populate
the chart with additional information - points, and in some cases,
portions of a trend line. This requires that we modify the original
JavaScript \ear{source} code to accommodate these additional elements,
which we generate using \texttt{r2d3}, as described above.

We defined JavaScript functions to draw the initial plot and set up
drawable points for the user drawn line. Drag events in \texttt{D3.js}
are utilized to observe and react to user input, as in the original `You
Draw It' script from the NYT. \footnote{The full source code we
  developed for the YouDrawIt task is available at
  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}

One constraint we inherited from the NYT code is that users can only
draw one-to-one functions; because the code works with drag events, each
point in \(x\) can correspond to only one point in \(y\), at least as
the code is currently written. While this is not particularly
problematic for our applications to date, it might limit applications of
this type of user interaction when assessing user drawn confidence
bands, ribbons, and other situations where two or more vertical points
are required.

@@

During testing of our modifications to the You Draw It' script, we
discovered that frequently the JavaScript code would ``skip'' from point
to point out of sequence. This resulted in a jagged line (visually) with
missing values in the underlying stored array of data; as a result, the
for-loop controlling the user-drawn line never exited and the user's
data was not recorded. While it would have been possible to fix this
using some sort of interpolation algorithm, we did not want to
compromise user results by introducing interpolation artifacts, so we
opted instead for a visual cue that fixed the problem from a
psychological angle.

@@

One challenge introduced by adding this visual cue is that D3 uses
Scalable Vector Graphics (SVG) to render elements. Adding an additional
layer meant that we had to ensure that not only the grids, lines, and
points of the default scatter plot and trend line rendered in the
correct order, but that the yellow box rendered behind all of these
points as well so that no information was masked.

@@

Shiny Messages are used to communicate the user interaction between the
JavaScript code and the R environment. The initial plot is rendered
using Shiny's \texttt{RenderD3} and \texttt{d3Output} functions;
subsequent user interactions are controlled via the JavaScript code.
Once the user is finished modifying the plot, they can submit their
response, so long as they have filled in all of the points. This is
enforced using Shiny's message-passing interface and a hook in the
JavaScript code that notifies Shiny when the array of user-drawn points
is completely filled in. Once the user is done drawing the line, we save
the results of the drawn line to a SQLite database, using Shiny to pass
data from JavaScript to R. The connections between each portion of the
app are shown in Figure~\ref{fig-you-draw-it-code-sketch}.

@@

\hypertarget{validation-study}{%
\subsection{Validation Study}\label{validation-study}}

We conducted a study in order to validate `You Draw It' as a method for
graphical testing, comparing results to the less technological method
utilized in Mosteller et al. (1981). The original study asked 153
graduate students and postdoctoral researchers in an Introductory
Biostatistics course to fit lines by eye to a set of four points (S, F,
V, N), each set with differing slope and variance properties, using an
8.5 x 11 inch transparency with a straight line etched across the
middle. Researchers conducted the study with a latin square experimental
design to test for a practice effect on the task. Results indicated
participants tended to fit the slope of the first principal component
(minimizes orthogonal residuals) as opposed to the slope from the
ordinary least squares regression (minimizes vertical residuals)
Figure~\ref{fig-pca-plot} and found no effect of order.

{\centering \includegraphics{./images/fig-pca-plot-1.pdf}

\caption{\label{fig-pca-plot}Comparison between an OLS regression
equation which minimizes the vertical distance of points from the line
and a regression equation with a slope calculated by the first principal
component which minimizes the smallest distance of points from the
line.}

In our study, we replicated Mosteller et al. (1981) using the `You Draw
It' method and simulated data with parameter coefficients selected to
reflect those from the four data sets in the original study. Data were
simulated based on a linear model with additive errors: \begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align} For each participant, a unique data set was randomly and
independently generated from the underlying parameters at the beginning
of the study and mapped to a scatter-plot graphic.
Figure~\ref{fig-eyefitting-simplot} illustrates an example of simulated
data for all four parameter choices intended to reflect the trends in
the original study. When participants started the study, they were first
asked to complete two practice plots - accompanied by instructions and a
.gif demonstrating the task - in order to train them in the skills
associated with executing the task. The four `You Draw It' task plots
associated with the selected parameters followed the practice plots in a
random order for each individual.

{\centering \includegraphics{./images/fig-eyefitting-simplot-1.pdf}

\caption{\label{fig-eyefitting-simplot}Example of simulated data points
displayed in a scatter-plot illustrating the trends associated with the
four selected parameter choices.}

Participants were recruited via Prolific in March 2022; while this study
does utilize a convenience sample, as this is primarily a perceptual
task, previous results have found few differences between expert and
non-expert participants in this context {[}CITE SPATIAL PAPER{]}. The
study was conducted and distributed via a Shiny application and
participants completed the tasks on their own computer, with a computer
mouse, and in an environment of their choosing. The data from this study
were collected to validate this method of graphical testing, with the
hopes of providing a new tool to assess graphical perception
interactively.

\hypertarget{data-analysis-and-results}{%
\subsection{Data Analysis and Results}\label{data-analysis-and-results}}

As a part of the validation study, we collected data for each
participant drawn trend line, allowing us to conduct formal analyses
which compared the intuitive visually fit lines to statistical
regression results - the original study lacked these formal comparisons,
providing only summaries of averages, variances, and actual (least
squares) values for the slope and intercept of each data set. The
feedback data, stored in a SQLite database, contained the simulated data
points\} - \((x_{ijk}, y_{ijk})\) -, the predicted values from the
statistical regression models - \(\hat y_{ijk,OLS}\), and
\(\hat y_{ijk,PCA}\) -, and the predicted values from the user drawn
line - \(y_{ijk,drawn}\) for parameter choice \(i = 1,2,3,4\),
\(j = 1,...N_\text{participant}\), and \(x_{ijk}\) value
\(k = 1, ...,4 x_{max} + 1\). Figure~\ref{fig-eyefitting-trial-plot}
displays an example of all three fitted trend lines for parameter choice
F.

{\centering \includegraphics{./images/fig-eyefitting-trial-plot-1.pdf}

\caption{\label{fig-eyefitting-trial-plot}Example of validation feedback
data where three trend lines show the the OLS fitted, PCA fitted, and
participant drawn values overlaid on the simulated data points.}

A unique data set was simulated independently for each participant,
therefore, comparisons of vertical residuals between the user drawn line
and the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were used to
assess the accuracy of participant drawn lines. We use mixed model
methods to analyze residual trends and capture the variability due to
participant differences.

We first fit a linear mixed model (LMM) to the OLS and PCA residuals
separately using the \texttt{lmer} function in the \texttt{lme4} package
(Bates et al. 2015), constraining the fit to a linear trend. Parameter
choice, \(x\), and the interaction between \(x\) and parameter choice
were treated as fixed effects with a random participant effect
accounting for variation due to participant. The LMM equation for each
fit (OLS and PCA) residuals is given by: \begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\)-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\)-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of \(x\)-value corresponding to either the OLS or
  PCA fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the \(x\)-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_\text{participant})\) is the random error
  due to the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

{\centering \includegraphics{./images/fig-eyefitting-lmer-residualplots-1.pdf}

\caption{\label{fig-eyefitting-lmer-residualplots}Estimated trends of
residuals (vertical deviation of participant drawn points from both the
OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed
model. A random sample of 75 participants was selected to display the
individual participant residuals behind the overall trend.}

Eliminating the linear trend constraint, we can extend the method and
analysis beyond ordinary least squares by using generalized additive
mixed models (GAMM) to estimate smoothing splines and allow for a more
flexible residual trend. The \texttt{bam} function in the \texttt{mgcv}
package was used to fit a GAMM separately to the OLS and PCA. Parameter
choice was treated as a fixed effect with no estimated intercept and a
separate smoothing spline for \(x\) was estimated for each parameter
choice. A random participant effect accounting for variation due to
participant and a random spline for each participant accounted for
variation in spline for each participant. The GAMM equation for each fit
(OLS and PCA) residuals is given by: \begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\)-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\)-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of \(x\)-value corresponding to either the OLS or
  PCA fit
\item
  \(\alpha_i\) is the intercept for the parameter choice \(i\)
\item
  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
\item
  \(x_{ijk}\) is the \(x\)-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_\text{participant})\) is the error due to
  participant variation
\item
  \(s_{j}\) is the random smoothing spline for each participant.
\end{itemize}
